Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
moving main discussion text to appendix, with short summary in main section adding notes

\subsection{Summary of Discussion}
Software engineering principles form the basis of methods, techniques, methodologies and tools. This group discussed the principles of software engineering design for sustainable software and their application in various domains including quantum chemistry and epidemiology.

\subsection{Description of Opportunity, Challenges, and Obstacles}
The opportunity was identified to distill existing software engineering and sustainability design knowledge into ''bite sized'' chunks for the Computational Science and Engineering Community. In addition, two primary challenges were identified:
\begin{itemize}
\item Mapping of the principles to best practice.
\item Demonstrating the return on investment of those best practices.
\end{itemize}

\subsection{Key Next Steps}

@@

\subsection{What Else is Needed?}
The following list of what else is required :

@@

\subsection{Description of Funding Needed}
Specific funding was not discussed in this working group. However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initatives.

>>>>

\subsection{Summary of Discussion}

The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy. Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion. There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community. Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems. It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members. The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles included: abstraction; coupling and cohesion; decomposition and modularization; encapsulation and information hiding; separation of interface and implementation; sufficiency completeness \& primitiveness; and separation of concerns. Similarly, user interface design principles included: learnability; user familiarity; consistency; minimal surprise; recoverability; user guidance; and user diversity. The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.

This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing. The group identified two example systems to discuss the application of the principles. The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids. The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets. The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference. This raised the question, what do the principles mean for  computational scientific and engineering software? Similarly, how do the principles relate to non-functional requirements? It was suggested that at the very minimum, that sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software:
\begin{itemize}
\item Extensibility: the software's ability to be extended and the level of effort required to implement the extension;
\item Maintainability: the effort required to locate and fix an error in operational software.
\end{itemize}
These fundamental building blocks could then be extended to include other quality attributes such as portability, reusability, scalability, usability, and energy efficiency etc. Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software. In addition, what do the five dimensions of sustainability mean for scientific software, i.e., environmental, economic, social, technical and individual?

\subsection{Description of Opportunity, Challenges, and Obstacles}
The opportunity was identified to distill existing software engineering and sustainability design knowledge into ''bite sized'' chunks for the Computational Science and Engineering Community. In addition, two primary challenges were identified:
\begin{itemize}
\item Mapping of the principles to best practices.
\item Demonstrating the return on investment of those best practices.
\end{itemize}

\subsection{Key Next Steps}
These was not determined at the workshop. \todo{is this correct, or should something go here?}

@@

\subsection{What Else is Needed?}

@@

\subsection{Description of Funding Needed}
Specific funding was not discussed in this working group. However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initiatives.

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor update
moving main discussion text to appendix, with short summary in main section adding notes

\subsubsection{Why it is important}

Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{}. However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{}. Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{}.

@@

The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy. Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion. There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community. Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems. It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members. The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles included: Abstraction; Coupling and cohesion; Decomposition and modularization; Encapsulation and information hiding; Separation of interface and implementation; Sufficiency completeness \& primitiveness; and Separation of concerns. Similalry, user interface design principles included: Learnability; User familiarity; Consistency; Minimal surprise; Recoverability; User guidance; and User diversity. The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}. The maifestio states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
%\todo{should these sets of bullet points be moved to the appendix? or moved into paragraph form, since they are relatively short items?}
A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.

This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing. The group identified two example systems to discuss the application of the principles. The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids. The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets. The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.

\subsubsection{Plans}
The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a set of guidelines on the principles and how they exemplary apply to scientific software system.

>>>>

%\subsubsection{Why it is important}

\todo{fill in cites in this paragraph}
Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{}. However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{}. Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{}.

@@

Software engineering principles form the basis of methods, techniques, methodologies and tools. This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}, Tate~\cite{tate2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.  The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements? It
appeared that the sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software: extensibility, the software's ability to be extended and the level of effort required to implement the extension; and
maintainability: the effort required to locate and fix an error in operational software.

For more information about the discussion, see Appendix~\ref{sec:appendix_eng_design}.

\subsubsection{Plans}
The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a set of guidelines on the principles and how they exemplary apply to scientific software system. Preliminary work will be carried out through undergraduate or post-graduate student projects.
